Factors Influencing Abundance & Virus Isolation in Mosquitoes

### 2024-04-24 products/manuscript/Manuscript.qmd
Updates to manuscript

**This project will lay the foundation for dissertation research that
will begin in June, 2024.**

@@

## Questions/Hypotheses to be addressed

*This project will lay the foundation for my dissertation research.*
**The goal for this phase of the project is to answer which
environmental, ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts.** From prior
research we know that the mosquito life cyle is influenced by a number
of factors. Temperature plays a major role in the mosquito life cyle
virus transmission dynamics. As temperature increases, the rate of
larval development also increases and the time between when a mosquito
ingests virus in a bloodmeal and when it can transmit it decreases. The
result is that the mosquito population can grow at faster rate and they
will have more time to transmit the virus. However, MEL larvae develop
in well insulated crypts that are less sensitive to ambient air
temperature changes, so the effect of temperature changes on MEL
development speed may be less pronounced than other vector species that
develop in open, standing water, like ditches and containers.
Precipitation increases the amount of standing water and increases the
chances of larval survival during the mosquito season and also during
the winter months for those species that overwinter in the larval stage.
For EEE, the main enzootic vector, MEL, overwinters as larvae.

While these relationships are well known, they are not well described
for EEE, especially in terms of quantifying the role these two factors
play in EEE transmission. Specifically, there's a gap in our
understanding of how these factors influence the number of mosquitoes
caught in traps and the rate of virus infection among those mosquitoes.
This gap is more pronounced when distinguishing the impacts on enzootic
(mosquitoes that primarily feed on birds) versus bridge vector
(mosquitoes that can transmit the virus from birds to humans) mosquito
species. Even less well described is the effect the factors have on
enzootic vs. bridge vector mosquito species.

@@

While the following questions will not be addressed in this project, it
will hopefully inform research that will answer the following questions
in dissertation research beginning this summer:

1\. Can an AI-driven model that leverages historic mosquito
surveillance, ecological, and environmental data accurately quantify the
risk of human EEEv infection in southeastern Massachusetts?

2\. What is the potential of machine learning algorithms to identify
early warning signals for EEE outbreaks, enabling timely public health
interventions in Massachusetts to prevent human infections?

3\. How can machine learning models use real-time data on mosquito
activity, environmental factors, and weather to accurately predict when
the risk of EEE transmission to humans is low enough to lift evening
outdoor activity restrictions? \*

>>>>

@@

## Questions/Hypotheses to be addressed
**Which environmental,ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts and which
models best describe the relationship between these factors?** 
From prior
research we know that the mosquito lifecyle is influenced by a number
of factors. Temperature plays a major role in the mosquito life cyle
virus transmission dynamics. As temperature increases, the rate of
larval development also increases and the time between when a mosquito
ingests virus in a bloodmeal and when it can transmit it decreases. The
result is that the mosquito population can grow at faster rate and they
will have more time to transmit the virus. However, MEL larvae develop
in well-insulated crypts that are less sensitive to ambient air
temperature changes, so the effect of temperature changes on MEL
development speed may be less pronounced than other vector species that
develop in open, standing water, like ditches and containers.
Precipitation increases the amount of standing water and increases the
chances of larval survival during the mosquito season and also during
the winter months for those species that overwinter in the larval stage.
For EEE, the main enzootic vector, MEL, overwinters as larvae.

While these relationships are well known, they are not well described
for EEE, especially in terms of quantifying the role these two factors
play in EEE transmission. Specifically, there's a gap in the
understanding of how these factors influence the number of EEE vectors mosquitoes
collected in traps and the rate of virus infection among those mosquitoes.

@@
